---
title: "PS 138Z - The Politics of Immigration: Section 7"
output:
  pdf_document:
  df_print: paged
  html_document: default
date: "2024-03-7"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Today, we will continue using `ggplot` to analyze data from the **Mexican Migration Project (MMP) Database** to produce.

# Cleaning data

First, let's load the data using the `read.csv()` function. This time, we will assign the dataset to an object called `mmp.data`.

```{r} 
mmp.data <- read.csv("mmp_subset.csv")
```

Let's upload `tidyverse` package only. You can upload a (pre-installed) package with the `library()` function.

```{r, message = FALSE}
library(tidyverse)
```

In this exercise, we will work with four variables: "Currently on last migration" (`uscurtrp`), "age" (`age`), "sex" (`sex`), and "education years" (`edyrs`). Remember that you can see the description of these and other variables in the dataset in the codebook. Let's start by creating a dataset with only these variables. Complete the code below: 

```{r}
no.miss.data <- mmp.data 
no.miss.data <- cbind(no.miss.data$uscurtrp, no.miss.data$age, 
                      no.miss.data$sex, no.miss.data$edyrs)
no.miss.data <- as.data.frame(no.miss.data )
names(no.miss.data) <- c("uscurtrp", "age", "sex", "edyrs")
```

Now, let's examine if any of these four variables has any numerical missing values. You can do this in multiple ways. This time, we will use the `range()` function. Complete the code below (Hint: remember that you should use the argument `na.rm = TRUE` because we do not know if the variables have missing values coded as `NA` as well): 


```{r}
range(no.miss.data$uscurtrp, na.rm = TRUE)
range(no.miss.data$age, na.rm = TRUE)
range(no.miss.data$sex, na.rm = TRUE)
range(no.miss.data$edyrs, na.rm = TRUE)
```

*What can you conclude?* *Which variables have numerical missing values?*

```{r}


```


Remember that for some functions, such as `mean()` and `range()`, we can ignore missing values coded as `NA` with the argument `na.rm = TRUE`. Also, remember that we can easily drop all the rows with at least one `NA` using the function `omit.na()`. For these reasons, we may prefer to replace the numerical missing values with `NA`s. Complete the code below:  


```{r}
no.miss.data$uscurtrp[no.miss.data$uscurtrp == 8888] <- NA
no.miss.data$age[no.miss.data$age == 8888] <- NA
no.miss.data$edyrs[no.miss.data$edyrs == 8888] <- NA
```

Now that we have replaced all the numerical missing values with `NA`s, we can drop any rows with `NA`s. 

```{r}
no.miss.data <- na.omit(no.miss.data)
```

To this point, we have a "clean" dataset with only the variables we are interested in and no missing values. As a final step, let's recode the `sex` and `uscurtrp` variables to facilitate plotting. Run the code below:  

```{r}
no.miss.data$uscurtrp[no.miss.data$uscurtrp == 1] <- "Yes"
no.miss.data$uscurtrp[no.miss.data$uscurtrp == 2] <- "No"

no.miss.data$sex[no.miss.data$sex == 1] <- "Male"
no.miss.data$sex[no.miss.data$sex == 2] <- "Female"
```

Use the `head()` function to look at the first 15 rows in the dataset.

```{r}
head(no.miss.data, 15)
```

# Characterizing recent immigrants

Complete the code below to calculate the average age and educational level by immigration status. 

```{r}
no.miss.data %>%
  group_by(uscurtrp) %>%
  summarise_at(vars(age, edyrs), c(mean), na.rm = TRUE)
```

Complete the code below to cross-tabulate immigration status and sex. 

```{r}
table(no.miss.data$sex, no.miss.data$uscurtrp)
```

*What can you conclude from these descriptive statistics?*


```{r}


```

Now, let's use the `ggplot` package to plot a grouped histogram showing the age distribution by immigration status. Complete the code below: 

```{r}
plot1 <- ggplot(no.miss.data, aes(x=age, color= factor(uscurtrp))) + 
  geom_histogram(binwidth = 2, fill="white", alpha=0.5) + 
  labs(
    title = "Age Distribution by Immigration Status",
    caption = "Source : Mexican Migration Project",
    x = "Age (in years)",
    y = "Frequency", 
    color = "Currently on last US migration") + 
  scale_color_manual(labels = c("Yes", "No"), values = c("darkblue", "darkred")) +
  theme_test()
plot1
```

Similarly, let's use the `ggplot` package to plot a grouped histogram showing the distribution of school years completed by immigration status. Complete the code below: 


```{r}
plot2 <- ggplot(no.miss.data, aes(x=edyrs, color= factor(uscurtrp))) + 
  geom_histogram(binwidth = 1, fill="white", alpha=0.5) + 
  labs(
    title = "Age Distribution by Immigration Status",
    caption = "Source : Mexican Migration Project",
    x = "Age (in years)",
    y = "Frequency", 
    color = "Currently on last US migration") + 
  scale_color_manual(labels = c("Yes", "No"), values = c("darkblue", "darkred")) +
  theme_test()
plot2
```

Finally, let's use `geom_bar()` to create a grouped bar plot showing how individuals with different immigration statuses are distributed by sex. Complete the code below:

```{r}
plot2 <- ggplot(no.miss.data, aes(x = uscurtrp, fill = factor(sex))) +
  geom_bar() +
  labs(
    title = "Sex Distribution by Immigration Status",
    caption = "Source : Mexican Migration Project",
    x = "Sex",
    y = "Frequency",
    legend = "Sex") +
  scale_fill_manual(values = c("darkgray", "darkolivegreen"),
                    name = "Sex") +
  theme_test()
plot2
```

*How do the descriptive tables and figures we produced relate to the theories we have discussed in class?*

```{r}


```
